Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews

### 2025-05-17 analysis/paper/paper.qmd
update readme

---
title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The status of archaeology as a science has been debated for decades and influences how we practice and teach archaeology. This study presents a novel bibliometric assessment of archaeology’s status relative to other fields using a hard/soft framework. It also presents a systematic review of computational reproducibility in published archaeological research. Reproducibility is a factor in the hardness/softness of a field because of it importance in establishing consensus. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A review of reproducibility assessments for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

@@

In their paper celebrating the 40th anniversary of this journal Torrence et al [-@torrenceFortyYearsStill2015] noted that reproducibility was an issue important to the reputation and sustainability of the discipline, and necessary for archaeological science to behave like a science. As part of the celebration of the 50th anniversary, and of Torrence's leadership of the journal, my contribution revisits these topics of archaeology's status as a science, this journal's place in the landscape of archaeological science, and how the journal has responded to a growing recognition of the importance of reproducibility. I first present bibliometric evidence of the position of archaeology as a whole, and this journal in particular, in the sciences. Next I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibilty of the research submitted for publication. Finally, I analyse twelve months of reproducibility reviews to identify common weaknesses in the ways archaeologists are working currently, and provide simple recommendations for researchers to overcome these and contribute to the improvement of computational reproducibility in archaeological science.

@@

What has been missing from these debates is a macroscopic observation of what the majority of archaeologists are actually doing, and an empirical comparison to a broad spectrum of relatively harder and softer disciplines. At the 'hard' end of the spectrum (e.g. physics and chemistry), scholars more typically share a large set of established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results. At the 'soft' end of the spectrum (e.g. economics and psychology), the set of theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new findings and the continuing relevance of previous work. In sum, the hard-soft status is defined by the amount of consensus in a field, and the speed at which consensus is reached on new knowledge [@fanelliBibliometricEvidenceHierarchy2013]. Hardness and softness is a controversial distinction, in part because it is sometimes used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983]. On the other hand, analyzing the characteristics that lead to the hard-soft distinction can be useful for understanding the diversity of academic inquiry, such as how different fields approach knowledge and differences in what counts as evidence and modes of argument, where fruitful collaborations might be possible due to shared methods and assumptions, and for curriculum design to structure courses appropriately based on a field's typical ways of knowing.

@@

To objectively quantify the diversity of modern archaeological practice across a scale of relative hardness or softness, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach. This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters. @fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources. In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, softer sciences include social sciences (Psychiatry, Psychology, Economics, Business and General Social Sciences), and the Humanities define the soft end of the spectrum. Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.

@@

While Fanelli and Glänzel (2013) analysed papers published in a single year (2012), I found only `r n_items_df_2012` papers for that same year, and `r round(prop_pub_after2012*100, 0)`% of papers in the sample published after that date. To make efficient use of the available data and ensure robust representation from different areas of archaeology, including those with lower frequencies of journal article publication, I analysed 9697 papers published during 1975-2025. This sample was collected from Clarivate’s Web of Science database by first selecting the Web of Science category ‘Archaeology’ and the Document type ‘article’ (n = 28,871). To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate’s Journal Citation Indicator. Finally I excluded journals with less than 100 articles in the database, resulting in `r n_journals` journals.

@@

@fig-compare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013. The most striking indicator of archaeology as a hard science is the number of authors, where it is between the social sciences and physics. Archaeology is a close fit with the social sciences in relative title length. It is between the social sciences and humanities in recency of references and diversity of references. The clearest indicator of archaeology as a soft science is article length where it is similar to the humanities. Overall archaeology does not sit squarely at either end of the hard-soft spectrum. It is generally not a harder science than the social sciences, with the exception of collaborator group sizes.

@@

# get this from the supplement

@@

The *Journal of Archaeological Research* is notable because it consistently ranks as soft, ranking as the softest journal for four of our five bibliometric variables. This is a predicable result for a review journal, which is a distinct type of journal dedicated to summarizing, analyzing, and synthesizing existing research in a particular field. The stated aim of the *Journal of Archaeological Research* is to 'bring together the most recent international research summaries on a broad range of topics and geographical areas' [@feinmanAimsScopeJournal2024]. A typical article is a long single-authored synthesis of archaeology in a region or on a topic. As the only review journal in this sample, this is a stark contrast to the other journals here that present original research findings, and like the *Journal of Cultural Heritage*, may be considered an outlier in this sample.

The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left. The variables that contribute to variation in PC1 are title length, number of pages, and the diversity of references. Journals with higher PC1 values have articles with longer titles, fewer pages, and less diverse reference lists. The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum. Variation in PC2 is influenced by the number of authors and recency of references. The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range of the PC2 axis (e.g. *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), characterised by fewer authors and more recent references. In the positive range of the PC2 axis are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g. *Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*). The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to both specialised and generalist communities of practice in archaeology.

@@

This macroscopic perspective derived from an analysis of the ways thousands of archaeologists communicate their research has produced a complex picture of archaeology as a science. In the context of a broad spectrum of other research areas, archaeologists behave like social scientists. We are harder than typical social scientists in tending to form larger groups of collaborators more often, and softer in sometimes writing longer articles than more resemble humanities scholarship. The outlook for the future of archaeology is also complex, with three out of five of the bibliometric variables trending towards more humanistic styles of working, but the discipline showing more extreme values in some metrics towards both hard and soft sciences after about 2010. Among archaeology journals, we see distinct communities of practice reflected in the PCA results that are very close together on the hard-soft spectrum, but have minor differences in their communication styles, perhaps due to differences in writing traditions inherited from parent disciplines such as geology and biology.

While these bibliometric variables provide several interesting insights into the status of archaeology as a science, via measurement of consensus, and are important for moving the debate beyond discussions of a small number of case studies, they miss a crucial factor that separates scientific practice from non-science. This is reproducibility, which, according to a report by the US National Science Foundation [@cacioppo2015social], "refers to the ability of a researcher to duplicate the results of a prior study using the same materials as were used by the original investigator. That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results... Reproducibility is a minimum necessary condition for a finding to be believable and informative." Scientific reproducibility is a factor contributing to the hardness of a field. Specifically, reproducibility is linked to to concept of consensus in a field because if more researchers provide sufficient detail for others to reproduce their results, then consensus on new knowledge can more often, and more, rapidly be established. The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan]. Boyle was concerned about the secrecy common among experimentalists in the 17th century and aimed to shift the culture from valuing direct in-person witnessing of scientific demonstrations towards meticulous written communications that were detailed enough to enable a reader to successfully undertake the experiment themselves, independent of the original author.

@@

In 2024 the *Journal of Archaeological Science* introduced a new kind of peer review that has provided an opportunity to tackle this question about computational reproducibility in archaeology. In January 2024 I accepted the position of 'Associate Editor for Reproducibility' (AER) for JAS and conducted reproducibility reviews of submissions that mentioned programming languages such as R or Python in the methods sections. A reproducibility review examines the code and data used to generate the results presented in the paper, and attempts to run the authors' code to reproduce the their results (see @editorsReproducibilityJournalArchaeological for more details about this process). This new AER role is based on similar positions (i.e. 'data editor' or 'reproducibility editor') that journals in economics [@vilhuber2019report], statistics [@wrobel2024partnering], astronomy [@muench2023roles], ecology [@bolnick2022ensuring], and environmental studies [@rosenbergReproducibleResultsPolicy2021] have had, in some cases for over a decade. In 2024, three archaeology journals, in addition to JAS, added AERs to their editorial communities: *Advances in Archaeological Practice* [@marwick2024introducing, one paper reviewed], *Journal of Field Archaeology* [@farahani2024reproducibility, two papers reviewed], and *American Antiquity* [@martin2024editors, no papers reviewed].

@@

```{r}
#| label: fig-aer-summary
#| fig-cap: "Summary of reproducibility reviews for JAS. A: Primary software used for the computational analysis reported in a manuscript. B: Computational or statistical method used by the authors (GMM = geometric morphometrics; Frequentist = hypothesis tests such as chi-square and ANOVA; AI/ML = artifical intellegence and machine learning, including neural networks and deep learning; MCMC = Markov Chain Monte Carlo,  i.e. Bayesian models and other simulations; Network = statistical analysis of social networks; 3D = analysis of 3D data such as artefact models; Composition = compositional analysis of artefacts). C: Locations where authors deposited their code and data files. D: Issues that prevented the reproducible review from succeeding on the first attempt. E: Relationship between software used and issues that make research irreproducible."
#| fig-height: 3

@@

At the time of writing (January 2025) we have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS (most papers required multiple reviews). Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by one measure, of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values available from a variety of hard and soft sciences.

@@

Despite the relatively small number of reproducibility reviews reported on here, there are patterns of common issues that point to a small set of simple tasks authors can do that have high potential to increase reproducibility. The problem of incomplete materials can be tackled in several basic ways. First, authors should use a simple and logical folder structure to organise their code and data to be as self-contained as possible (ie. not requiring files outside of the project folder or any modifications by a reader to run the code). There are many excellent, simple, and widely-used templates that authors can choose from, e.g. @doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017

Second, for analyses that are not highly time-consuming (which was over 90% of the sample), authors should re-run their code more than once, and ideally not on the same computer (e.g. by another co-author of the paper), before submission to confirm everything works as expected. This would ensure portability and likely also prevent issues relating to path and function errors. Complex and time-consuming analyses should use pipeline or workflow management tools, e.g. GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code by others [@wrattenReproducibleScalableShareable2021; @landau2021targets].

Third, authors should include in their compendium a README document that describes to readers the folders and files contained in the project. The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g. what order to run the code files in). A README should also briefly describe the contents of the compendium, where other necessary files can be obtained (e.g. data files not included in the compendium), the key software packages needed and the version numbers that the authors used, and if the analysis takes more than a few minutes to run on a typical laptop, the hardware resources and compute time used by the author.

Fourth, authors should ensure clear, direct and obvious connections between their code and the results they present in their paper. One simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption. Another way some authors are accomplishing this is by using literate programming tools, such as Quarto and Jupyter notebooks [@quarto; @kluyver2016jupyter] that enable the research narrative and code for data analysis to be woven together in one document. Quarto was a popular tool among the JAS papers in the reproducibility review sample, for example, @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024 wrote their entire manuscripts using Quarto. These four simple steps have high potential to substantially increase the computational reproducibility of archaeological science.

Our finding that dependencies are a common cause of irreproducible results is consistent with previous studies that have identified this as a widespread weakness in communicating computationally intensive research [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024]. In our sample, issues relating to dependencies are strongly associated with the use of Python. One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020]. Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse and has evolved practices to effectively communicate dependencies [e.g. @bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].

@@

>>>>

---
title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The status of archaeology as a science has been debated for decades and influences how we practice and teach archaeology. This study presents a novel bibliometric assessment of archaeology’s status relative to other fields using a hard/soft framework. It also presents a systematic review of computational reproducibility in published archaeological research. Reproducibility is a factor in the hardness/softness of a field because of its importance in establishing consensus. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A review of reproducibility assessments for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

@@

In their paper celebrating the 40th anniversary of this journal Torrence et al. [-@torrenceFortyYearsStill2015] noted that reproducibility was an issue important to the reputation and sustainability of the discipline, and necessary for archaeological science to behave like a science. As part of the celebration of the 50th anniversary, and of Torrence's leadership of the journal, my contribution revisits these topics of archaeology's status as a science, this journal's place in the landscape of archaeological science, and how the journal has responded to a growing recognition of the importance of reproducibility. I first present bibliometric evidence of the position of archaeology as a whole, and this journal in particular, in the sciences. Next, I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibility of the research submitted for publication. Finally, I analyse twelve months of reproducibility reviews to identify common weaknesses in the ways archaeologists are working currently, and provide simple recommendations for researchers to overcome these and contribute to the improvement of computational reproducibility in archaeological science.

@@

What has been missing from these debates is a macroscopic observation of what the majority of archaeologists are actually doing, and an empirical comparison to a broad spectrum of relatively harder and softer disciplines. At the 'hard' end of the spectrum (e.g. physics and chemistry), scholars more typically share a large set of established set of theories, facts, and methods, facilitating fairly rapid agreement on the validity and significance of new results [@biglan1973characteristics]. At the 'soft' end of the spectrum (e.g. economics and psychology), the set of theories, facts, and methods on which there is widespread consensus is smaller, and agreement is slower and less frequently reached about the significance of new findings and the continuing relevance of previous work. In sum, the hard-soft status is defined by the amount of consensus in a field, and the speed at which consensus is reached on new knowledge [@fanelliBibliometricEvidenceHierarchy2013]. Hardness and softness is a controversial distinction, in part because it is sometimes used to imply a rank order of disciplines that encodes legitimacy, productivity, perceived value to society, and worthiness of funding [@DifferentAgenda2012; @coleHierarchySciences1983]. Another criticism is that it may be more of an emergent product of social and institutional processes rather than intrinsic differences in method or consensus [@latour1987science]. On the other hand, analyzing the characteristics that lead to the hard-soft distinction can be useful for understanding the diversity of academic inquiry, such as how different fields approach knowledge and differences in what counts as evidence and modes of argument, where fruitful collaborations might be possible due to shared methods and assumptions, and for curriculum design to structure courses appropriately based on a field's typical ways of knowing [@becher2001academic].

@@

To objectively quantify the diversity of modern archaeological practice across a scale of relative hardness or softness, as an evaluation of its status as a science, and the place of this journal in context of other archaeology journals, I take a bibliometric approach. This approach is based on @fanelliBibliometricEvidenceHierarchy2013, who examined the hardness and softness of 12 disciplines using scholarly publication parameters. @fanelliBibliometricEvidenceHierarchy2013 found a spectrum of statistically significant variation in bibliometric variables from the physical to the social sciences, with papers tending at the softer end of the spectrum tending to have fewer co-authors, use less substantive titles, have longer texts, cite older literature, and have a higher diversity of sources. In @fanelliBibliometricEvidenceHierarchy2013's analysis harder sciences include Space Science, Physics, Chemistry, softer sciences include social sciences (Psychiatry, Psychology, Economics, Business, and General Social Sciences), and the Humanities define the soft end of the spectrum. Following @fanelliBibliometricEvidenceHierarchy2013, I quantify the number of authors, length of article, relative title length, age of references, and diversity of references for a large sample of peer-reviewed journal articles.

@@

While Fanelli and Glänzel (2013) analysed papers published in a single year (2012), I found only `r n_items_df_2012` papers for that same year, and `r round(prop_pub_after2012*100, 0)`% of papers in the sample published after that date. To make efficient use of the available data and ensure robust representation from different areas of archaeology, including those with lower frequencies of journal article publication, I analysed 9697 papers published during 1975-2025. This sample was collected from Clarivate’s Web of Science database by first selecting the Web of Science category ‘Archaeology’ and the Document type ‘article’ (n = 28,871). To focus on journals of broad relevance to most archaeologists, and that are representative of substantial communities of practice, I then filtered the results to keep only articles published in the top-ranking 25 journals according to their h-indices as reported by Clarivate’s Journal Citation Indicator. Finally, I excluded journals with less than 100 articles in the database, resulting in `r n_journals` journals.

@@

@fig-compare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013. The most striking indicator of archaeology as a hard science is the number of authors, where it is between the social sciences and physics. Archaeology is a close fit with the social sciences in relative title length. It is between the social sciences and humanities in recency of references and diversity of references. The clearest indicator of archaeology as a soft science is article length where it is similar to the humanities. Overall, archaeology does not sit squarely at either end of the hard-soft spectrum. It is generally not a harder science than the social sciences, with the exception of collaborator group sizes.

@@

# get this from the supplement on GAMS
knitr::include_graphics(here::here("analysis/figures/fig-smooth-plots-paper.png"))

@@

The *Journal of Archaeological Research* is notable for consistently ranking as soft; it was the softest journal for four of our five bibliometric variables. This is a predicable result for a review journal, which is a distinct type of journal dedicated to summarizing, analyzing, and synthesizing existing research in a particular field. The stated aim of the *Journal of Archaeological Research* is to 'bring together the most recent international research summaries on a broad range of topics and geographical areas' [@feinmanAimsScopeJournal2024]. A typical article is a long single-authored synthesis of archaeology in a region or on a topic. As the only review journal in this sample, this is a stark contrast to the other journals here that present original research findings, and like the *Journal of Cultural Heritage*, may be considered an outlier in this sample.

The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left. The variables that contribute to variation in PC1 are title length, number of pages, and the diversity of references. Journals with higher PC1 values have articles with longer titles, fewer pages, and less diverse reference lists. The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum. Variation in PC2 is influenced by the number of authors and recency of references. The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range of the PC2 axis (e.g. *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), characterised by fewer authors and more recent references. In the positive range of the PC2 axis are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g. *Environmental Archaeology*, *Geoarchaeology*, *Archaeological Research in Asia*, *Journal of Island and Coastal Archaeology*). The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to both specialised and generalist communities of practice in archaeology.

@@

This macroscopic perspective derived from an analysis of the ways thousands of archaeologists communicate their research has produced a complex picture of archaeology as a science. In the context of a broad spectrum of other research areas, archaeologists behave like social scientists. We are harder than typical social scientists in tending to form larger groups of collaborators more often, and softer in sometimes writing longer articles than more resemble humanities scholarship. The outlook for the future of archaeology is also complex, with three out of five of the bibliometric variables trending towards more humanistic styles of working, but the discipline showing more extreme values in some metrics towards both hard and soft sciences after about 2010. Among archaeology journals, we see distinct communities of practice reflected in the PCA results that are very close together on the hard-soft spectrum, but have minor differences in their communication styles, perhaps due to cultural differences in writing traditions inherited from parent disciplines such as geology and biology [@becher2001academic].

While these bibliometric variables provide several interesting insights into the status of archaeology as a science, via measurement of consensus, and are important for moving the debate beyond discussions of a small number of case studies, they miss a crucial factor that separates scientific practice from non-science. This is reproducibility, which, according to a report for the US National Science Foundation [@cacioppo2015social], "refers to the ability of a researcher to duplicate the results of a prior study using the same materials as were used by the original investigator. That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results... Reproducibility is a minimum necessary condition for a finding to be believable and informative." Scientific reproducibility is a factor contributing to the hardness of a field. Specifically, reproducibility is linked to to concept of consensus in a field because if more researchers provide sufficient detail for others to reproduce their results, then consensus on new knowledge can more often, and more, rapidly be established. The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan]. Boyle was concerned about the secrecy common among experimentalists in the 17th century and aimed to shift the culture from valuing direct in-person witnessing of scientific demonstrations towards meticulous written communications that were detailed enough to enable a reader to successfully undertake the experiment themselves, independent of the original author.

@@

In 2024 the *Journal of Archaeological Science* introduced a new kind of peer review that has provided an opportunity to tackle this question about computational reproducibility in archaeology. In January 2024 I accepted the position of 'Associate Editor for Reproducibility' (AER) for JAS and conducted reproducibility reviews of submissions that mentioned programming languages such as R or Python in the methods sections, taking guidance from similar initiatives in other fields [e.g. @ivimey-cookImplementingCodeReview2023; @nustCODECHECKOpenScience2021]. A reproducibility review examines the code and data used to generate the results presented in the paper, and attempts to run the authors' code to reproduce their results (see @editorsReproducibilityJournalArchaeological for more details about this process). This new AER role is based on similar positions (i.e. 'data editor' or 'reproducibility editor') that journals in economics [@vilhuber2019report], statistics [@wrobel2024partnering], astronomy [@muench2023roles], ecology [@bolnick2022ensuring], and environmental studies [@rosenbergReproducibleResultsPolicy2021] have had, in some cases for over a decade. In 2024, three archaeology journals, in addition to JAS, added AERs to their editorial communities: *Advances in Archaeological Practice* [@marwick2024introducing, one paper reviewed], *Journal of Field Archaeology* [@farahani2024reproducibility, two papers reviewed], and *American Antiquity* [@martin2024editors, no papers reviewed at the time of writing].

@@

```{r}
#| label: fig-aer-summary
#| fig-cap: "Summary of reproducibility reviews for JAS. A: Primary software used for the computational analysis reported in a manuscript. B: Computational or statistical method used by the authors (GMM = geometric morphometrics; Frequentist = hypothesis tests such as chi-square and ANOVA; AI/ML = artificial intelligence and machine learning, including neural networks and deep learning; MCMC = Markov Chain Monte Carlo,  i.e. Bayesian models and other simulations; Network = statistical analysis of social networks; 3D = analysis of 3D data such as artefact models; Composition = compositional analysis of artefacts). C: Locations where authors deposited their code and data files. D: Issues that prevented the reproducible review from succeeding on the first attempt. E: Relationship between software used and issues that make research irreproducible."
#| fig-height: 3

@@

At the time of writing (January 2025) we have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS (most papers required multiple reviews). Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by one measure, of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97) offers another perspective. These 97 articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.). Under this broader scope, the success rate is 7%, a result also found in a study of 497 papers in 9 ecology journals [@kellnerFunctionalCodeRare2025]. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values available from a variety of hard and soft sciences.

@@

Despite the relatively small number of reproducibility reviews reported on here, there are patterns of common issues that point to a small set of simple tasks authors can do that have high potential to increase reproducibility. The problem of incomplete materials can be tackled in several basic ways. First, authors should use a simple and logical folder structure to organise their code and data to be as self-contained as possible. Authors should provide their materials organised such that a reader can successfully run all code as-is, without making any manual modifications (e.g. use relative rather than absolute file paths so that readers don't have to rename or move files around to make the code work) [@sandveTenSimpleRules2013]. Code and data files should be in the simplest format possible, for example a plain text R script file is smaller and easier to use than a PDF or Word document that includes R code. Script files should have the order in which they are to be run explicit in the file name, e.g. 001-load-data.R, 002-clean-data.R, 003-analyse-data.R. There are many excellent, simple, and widely-used project templates that authors can choose from that make it easy for authors to follow best practices of project organisation, e.g. @doi:10.1080/00031305.2017.1375986, @figueiredoSimpleKitUse2022, @Cookiecutter, @cooperGuideReproducibleCode2017 and @wilsonGoodEnoughPractices2017

Second, authors should include in their compendium a README document that describes to readers the folders and files contained in the project [@abdillHowtoGuideCode2024]. The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g. what order to run the code files in). A README should also briefly describe the contents of the compendium, where other necessary files can be obtained (e.g. data files that cannot be included in the compendium due for ethical or other reasons), the key software packages needed and the version numbers that the authors used, and if the analysis takes more than a few minutes to run on a typical laptop, the hardware resources and compute time used by the author.

Related to the basic documentation provided by the README, authors should document clear, direct and obvious connections between their code and the results they present in their paper [@sandveTenSimpleRules2013]. One simple way to do this is to have one code file for each figure and table, and name the code files with the figure or table number and some key words in the caption. Another way some authors are accomplishing this is by using literate programming tools, such as Quarto and Jupyter notebooks [@quarto; @kluyver2016jupyter] that enable the research narrative and code for data analysis to be woven together in one document. Quarto was a popular tool among the JAS papers in the reproducibility review sample, for example, @vernonMethodDefiningDispersed2024 and @ragnoSheepGoatsTaxonomic2024 wrote their entire manuscripts using Quarto.

Documentation is also a key tool is tackling the problems with dependencies described in the previous section. Our finding that dependencies are a common cause of irreproducible results is consistent with previous studies that have identified this as a widespread weakness in communicating computationally intensive research [@trisovicLargescaleStudyResearch2022; @samuelComputationalReproducibilityJupyter2024]. In our sample, issues relating to dependencies are strongly associated with the use of Python. One possible reason for this is that relative to R, Python uses more package managers, more environments, and deeper dependency chains with more complex inter-dependencies that change more rapidly [@decan2016topology; @decanEmpiricalComparisonDependency2019; @korkmazModelingImpactPython2020]. Another reason may be that there is a bigger and more established community of R users in archaeology [@schmidtToolDrivenRevolutionsArchaeological2020; @batistOpenArchaeologyOpen2024] that highly values code that is easy for others to reuse and has evolved practices to effectively communicate dependencies [e.g. @bilottiPointPatternAnalysis2024; @willExploringUtilityUnretouched2025].

@@

Finally, for analyses that are not highly time-consuming (which was over 90% of the sample), authors should re-run their code more than once, and ideally not on the same computer (e.g. by another co-author of the paper), before submission to confirm everything works as expected [@abdillHowtoGuideCode2024; @rothTenPrinciplesReliable2025]. This ensures portability and likely also prevents issues relating to path and function errors. Complex and time-consuming analyses should use pipeline or workflow management tools, e.g. GNU Make, Luigi, Snakemake, or Targets, to document the relationship of the files and folders in a machine-readable format and simplify running and re-running code by others [@wrattenReproducibleScalableShareable2021; @landau2021targets].

@fig-checklist summarises the key recommendations discussed in this section in a format that can be used as a check-list for authors submitting research for publication in JAS and other journals that do reproducibility reviews. This checklist is based on both the results presented in @fig-aer-summary and similar lists used by other journals, for example by the *Biometrical Journal* [@hornungBiometricalJournalChecklist] and *The Review of Financial Studies* [@perignonComputationalReproducibilityFinance2024].

```{r}
#| label: fig-checklist
#| fig-cap: "Checklist summarising a small set of some of the simplest tasks authors can do that have high potential to increase reproducibility"
#| fig-height: 4

# I made this using Canva, original is here: https://www.canva.com/design/DAGnj7LkvvE/lrjyzV1WywPWMDdo9nwJJg/edit 
knitr::include_graphics(here::here("analysis/figures/simple-reproducibility-checklist.png"))
```
